---
title: "GDA_MRK3"
author: '2020209'
date: "26/04/2022"
output: pdf_document
---
# Introduction
We are going to explore the factors which affect the distribution of Cotoneaster in the South West region of England. The Cotoneaster is an invasive species and we want to know what is effecting its growth in order to be able to control its spread.  
We are going to explore the affects temperature, rainfall and the proximity to differnet has on the distribution of the plant. 


 
```{r, echo= FALSE,message= FALSE, warning=FALSE}
 setwd("E:/GDA")
library("mosaic")
library("sf")
library("tmap")
library("tmaptools")
library("viridis")

COTO= read.csv("Cotoneaster_data.csv", header= TRUE)
```

# Data description 
To begin we will look at how the distribution varies by species:
```{r}
my_palette=c("Blue","orange","red","Green", "pink") 
my_settings = list(superpose.symbol = list(pch = 21,col= "black",fill = my_palette))
xyplot(Northing~Easting,
       main="Map of Cotoneaster distribution (Species)", 
       groups=species,
       data=COTO,
       par.settings=my_settings,
        ylab= "Northing", xlab="Easting",
       auto.key=list(space= "right", title= "Species", columns = 1))
```
We can see that there are some pockets towards the south west of Cornwall and around Bristol.
It also becomes clear that the majority of the data is two of the species; Cotoneaster Horizontalis and Simonisii. 

The variables we have access to are temperature, rainfall, and proximity to rock types. In addition to the location they are found in. 

Using the summary function, we can see the mean, median and quartiles of our variables where the plants are found. 

```{r}
summary(COTO [,c(3,4,5,6,7)])
```
This table shows lots of information but the most useful data comes from the temperature and rainfall results.  
The temperature where the plant is found varies between quite a small range of 8.3 degrees to 10.9 degrees. 
This small range suggests that the plant can only survive in these conditions so I am going to look into the temperature distribution.   

```{r}
hist(COTO$temperature_C, main = "Cotoneaster temperature data observations",
     col="lightblue", xlab="Temperature Degrees C")
```
For the locations where the Cotoneaster is found it would appear there is a normal distribution. With the plant being most suited to 9.6 degrees. 


# Summary Statistics 
To test if the temperature data is normally distributed we can use the Shapiro-wilk test. 
```{r}
shapiro.test(COTO$temperature_C)
```
The W value is very close to 1 which suggests a the data is normally distributed. 
However the p-value disagrees with this; as 1.522e-08 is less than the critical value for the 95% confidence level of 0.05 we reject the null hypothesis that the data is normally distributed. 
The data has a normal shape but is not normally distributed. 


# Inferrential statistics 
Parametric tests are preferred for inferential analysis as they have more statistical power. 
However, in order to perform parametric analysis a normal distribution is required. 
For temperature, the data is not normally distributed so we have to perform non-parametric tests.
The non-parametric tests consider similarities in the distributions based on the medians. 

We are going to use the Kruskal-Wallis test to analyze if there are differences between climatic conditions for Cotoneaster species. 

```{r}
kruskal.test(temperature_C~ species, data= COTO)
qchisq(p=0.05, df= 4, lower.tail= FALSE)
```
The p value is less than the 95% significance level (9.487729) meaning that there are differences in the medians of temperatures for the species. 
Cotoneaster Horiznatalis and Simonisii therefore react differently to temperature. 

# Regression analysis  
The next step is to look at other factors which affect the Cotoneaster
Lets have a look into how temperature and rainfall affect Cotoneaster 
We want to know if other environmental factors also influence this spread we are going to look at temperature and rainfall.

I am going to subset the data to compare differences between species. 
We want to use the most data possible so I am going to subset the largest species sets.
Fav stats shows the number of observations from each species.
```{r, echo=FALSE, eval= FALSE}
favstats(~temperature_C|species, data=COTO)
```

Cotoneaster Horizontalis and Cotoneaster Simonisii have the largest number of observations with 510 and 276 respectively, the larger the data set the more accurate our statistical analysis is going to be. The other species are significantly smaller, with n= 13, 45 and 6. 
So we will subset the data to use them.

```{r}
C_HORI= subset(COTO, species == "Cotoneaster horizontalis")
C_SIMO= subset(COTO, species == "Cotoneaster simonsii")
```

The first regression model will use rainfall and temperature in combination to predict the distribution of Horizontalias 
```{r}
cor.test(C_HORI$rainfall_mm, C_HORI$temperature_C)
```

The results show a correlation of 0.448 therefore there is a relationship between rainfall and temperature. This is not ideal for our regression analysis as they may influence each other but it is a low enough value to continue as it is below 0.5. 


The first regression model is going to look at the influence of temperature and rainfall for the Horizontalis 
```{r, echo=FALSE, eval= FALSE}
summary(lm(temperature_C~rainfall_mm, data= C_HORI))
```

The multiple R squared value of 0.2013 is low but it accounts for 20% accuracy in the model. 
There is room to improve this model; so we will have a look at other factors. 

```{r}
xyplot(temperature_C ~rainfall_mm, data= C_HORI ,type= c("p","r"), 
      xlab= "Rainfall (mm)", ylab= "Temperature (C)", 
      main= "Rainfall vs Temperature for Horizontalis ")
```
This plot shows that there is a weak positive correlation in Horizalias distribution between increased rainfall and increased temperature. 
This species of Cotoneaster prefers the warmer and wetter conditions. 

Then we are going to have a look at the Simonisii species. 

```{r, echo=FALSE, eval= FALSE}
cor.test(C_SIMO$rainfall_mm, C_SIMO$temperature_C)
summary(lm(temperature_C~rainfall_mm, data= C_SIMO))
```
For Simonisii there is a negative relationship between temperature and rainfall, so It tends to be more common in areas with higher temperatures but lower rainfall.

```{r}
xyplot(temperature_C ~rainfall_mm, data= C_SIMO ,type= c("p","r"), 
       xlab= "Rainfall (mm)", ylab= "Temperature (C)", 
       main= "Rainfall vs Temperature for Simo ")
```
The distribution for Simonisii is different from Horizontalis as prefers drier conditions. 


Let's have a look at another factor. what rocks does each species like to be near?

Using the regression model we had already made for the Horizontalis we are going to add chalk proximity as a factor to see if it increases the accuracy of the regression model. 

```{r, echo=FALSE, eval= FALSE}
summary(lm(temperature_C~rainfall_mm & Chalk_proximity, data= C_HORI))
```

When chalk proximity is included in the regression model the r squared decreases from 0.2013 to 0.0159  meaning the model becomes less effective. 
This means that the proximity to chalk is unlikely to be affecting the distribution of horizontals. 

Does the same thing happen for Simonisii? 
```{r, echo=FALSE, eval= FALSE}
summary(lm(temperature_C~rainfall_mm & Limestone_proximity, data= C_SIMO))
```
This also reduces the effectiveness of the model from 0.03783 to 0.0001805. 
Suggesting that the limestone proximity is not effective in predicting the distribution of Simonisii. 

For both species samples, the proximity to rock types makes the regression model less effective. 
We can conclude that temperature and rainfall are influencing the distribution of Cotoneaster but proximity to rock types does not. 
 

To further investigate the influence rock types have on Cotoneaster distribution We will make a map showing which samples are close to the underlying lithology. 

# Mapping 
```{r}
HORIZ_MAP = st_as_sf(C_HORI, coords= c("Easting", "Northing"))
st_crs (HORIZ_MAP)=27700
basemap <- read_osm(HORIZ_MAP, ext = 1.1)
tm_shape(basemap)+
  tm_rgb(alpha = 0.5)+
  tm_shape(HORIZ_MAP)+
  tm_dots("Chalk_proximity", palette= rev(turbo(100)), 
  size= 0.2, n=10, title= " Distance in m from chalk  ")+
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "bottom"))+
  tm_layout(bg.color = "white", 
  main.title = "Cotoneaster Horizontalis distance from Chalk",
  main.title.position = "centre")
```
As the regression model predicted, there is little correlation between Chalk proximity and Cotoneaster distribution. There are pockets in Cornwall which are around 200Km away from chalk and some that are as close as 2km.

We are going to do the same with the Simonisii results too.
```{r}
SIMO_MAP = st_as_sf(C_SIMO, coords= c("Easting", "Northing"))
st_crs (SIMO_MAP)=27700
basemap <- read_osm(SIMO_MAP, ext = 1.1)
tm_shape(basemap)+
  tm_rgb(alpha = 0.5)+
  tm_shape(SIMO_MAP)+
  tm_dots("Limestone_proximity", palette= rev(turbo(100)),
  size= 0.2, n=10, title= " Distance in m from Limestone")+
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "bottom"))+
  tm_layout(bg.color = "white", main.title = "Cotoneaster Simosii distance from Limestone"
  , main.title.position = "centre")
```
Again there are similar results as expected from the regression model; little correlation between the distribution of Simonisii and underlying limestone lithology. 

# Conclusion 

The distribution of Cotoneaster is dependent on the species, with factors having different influences on each other. When analyzing the most common species, we found that temperature and rainfall had significant influence on where the plant was distributed. With Cotoneaster Horizontalis preferring relatively warmer and wetter environments, and Simonisii being more suited to drier and warmer climates. Additionally, the influence of underlying lithology was ruled out as for both of the species investigated, little correlation was found. 
Our findings may be useful to controlling the spread of the invasive species as a greater understanding of the favored conditions are known. 
Further research into factors which can be controlled may be useful as changing the climate is not a possibility in controlling the spread of the species. 


